Two applications of logic programming to Coq

### 2020-11-06 papers/types2020/sec3.tex
Staging things to do a merge.

%\todo{Right now there is no intro to FPC}
% \begin{metanote}
%   (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). % In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
% \end{metanote}

@@

% The proof search approach to encoding Horn clause computation
% results in the structuring of proofs with repeated
% switching between a \emph{goal-reduction} phase and a
% \emph{backchaining} phase~\cite{miller91apal}.

%\subsection{Certificate checking with expert predicates}
%\subsection{Adding certificates and experts to a proof system}
%\label{ssec:fpc}

% \todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}

A \emph{Horn clause} is a formula of the form $\forall \bar
x_1. A_1\supset \forall \bar x_2. A_2\supset \forall \bar
x_n. A_n\supset A_0$ where $\forall\bar x_i$ denote a list of
universal quantifiers ($i\in\{1,\ldots,n\}$) and $A_0,\ldots,A_n$ are
atomic formulas.  It is well known that the following set of sequent
calculus proof rules are complete for both classical and
intuitionistic logic when one is attempting to prove that a given
atomic formula $A$ is provable from a give set $\Pscr$ of Horn
clauses. 
\[
  \infer[decide]{\Pscr\vdash A}{D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
  \qquad
  \infer[init]{\Pscr\Downarrow A\vdash A}{}
\]
\[
  \infer[\forall L]{\Pscr\Downarrow\forall x. D\vdash A}
        {\Pscr\Downarrow D[t/x]\vdash A}
  \qquad
  \infer[\supset L]{\Pscr\Downarrow B\supset D\vdash A}
        {\Pscr\vdash B\qquad \Pscr\Downarrow D\vdash A}
\]
Here, we use two different styles of sequents.  The sequent
$\Pscr\vdash A$ is the usual sequent which we generally use as the end
sequent (the conclusion) of a proof.  The sequent $\Pscr\Downarrow
D\vdash A$ is a \emph{focused} sequent in which the formula $D$ is the
\emph{focus} of the sequent.  The two left introduction rules and the
initial rule can only be applied to the focused formula.  This latter
point is in constrast to Gentzen's sequent calculus where these rules
can involve \emph{any} formula on the left of the $\vdash$.  The fact
that this proof system is complete for both classical and
intuitionistic logic (when restricted to the Horn clause fragment)
follows from rather simple considerations of Horn clauses
\cite{nadathur90jacm} and from the completeness of LJT proofs
\cite{Herbelin94}.  The use of the term \emph{focus} comes from
Andreoli's proof system for linear logic \cite{andreoli92jlc}.

@@

Figure~\ref{fig:augmented} contains these proof rules and extends it
in the following fours ways so that it more appropriately fits into
the Coq proof system.
\begin{enumerate}
  \item Instead of having separate connectives for $\forall$ and
    $\supset$, we have the dependent product connective $(x:A)D$. 
  \item We have incorporated both \emph{proof certificates} (using the
    schematic variable $\XXi$) along with expert predicates
    (predicates with the $e$ subscript).
  \item The inference rules are also anotated by terms structures that
    can be give directly to the Coq kernel for checking.
  \item We have added various premises which are responsible to
    pulling information across the Elpi API with Coq.
\end{enumerate}
The proof system described in Figure~\ref{fig:kernel} corresponds to
the Calculus of Inductive Constructions restricted to the Horn clause
fragment of pure \lsti{Inductive} definitions.
This proof system is inspired by the calculus for proof search in
Pure Type Systems introduced in \cite{LengrandDM06}, based in turn on
ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a
term language that includes terms and lists of terms, and two distinct
typing judgments for the two categories. In our case, where proof
search steps are viewed as computation, the term judgment $\XXi\vdash
t \colon A$ corresponds to a goal-reduction step while the judgment
$\XXi \Downarrow l \colon D\vdash A$ corresponds to a backchaining
step. This coincides with the idea behind the \emph{spine
  calculus}~\cite{Cervesato97tr}.  The main novelity of our proof
system here is that proof terms and proof certificates are used
simultaneously in all inference rules.

@@

The \emph{unfold} rule is the equivalent in our simplified setting of
a \emph{decide} rule in an explicitly focused system. From a logic
programming viewpoint, given an atom, the rule selects a clause on
which to back-chain on: we look-up the costructors from an
\lsti{Inductive} definition from the global enviroment, one that
matches the atom we wish to bachchain on and then call the latter
judgment. The rules for backchaining are standard, corresponding to
the \emph{init} and $\forall$-left rules.

@@

In figure~\ref{fig:augmented}  each
inference rule is augmented with an additional premise involving an
\emph{expert predicate}, a certificate $\Xi$, and possibly continuations of
certificates ($\Xi'$, $\Xi_1$, $\Xi_2$) with extracted information from
These notions are taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
\todo{MM: Talk about randomized FPC?}
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:augmented} and~\ref{fig:kernel}:
that is, the exact set of constructors for the  type of certificates as
well as the specification of the expert predicates listed \emph{ibidem}.
%
The top of Figure~\ref{fig:resources} displays two FPCs,
both of which can be used to describe proofs where we bound
the dimension of  a proof.
%
For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) B)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC can be used to bound the total number of instances of
unfoldings in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.

@@

Thanks to the Coq-Elpi interface, in particular to the ``main''
proocedure \lsti{solve}, we can package the \lP's code for the checker
as a tactic that can be called as any other tactic in a Coq script.

@@

The glue code between Coq-Elpi and the implementation of our calculus is
straightforward: the goal is given to us as a quadruple of a context, an
evar, a type and a list of extra information. In additon, the certificate is
supplied by the user when calling the tactic (it simply is an integer in this
case). We just need to call \lsti{check} on the goal's type, together with
the certificate, in order to obtain a reconstructed proof term. We avoid
calling the reconstruction directly on the evar because Coq-Elpi ensures that
evars manipulated by \lP are well-typed at all times, and we cannot guarantee
that since we work with partially reconstructed term; getting around this is
simply a matter of assigning the reconstructed term to the evar right
afterwards.
The following example shows how we can use the above tactic to do
FPC-driven logic programming in Coq and return a Coq proof-term:
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
eexists.
elpi  dprolog height 10.
Qed.
Print i1.
ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)) : exists R : list nat, insert 2 [0; 1] R
\end{lstlisting}

The \lsti{dprolog} tactic can be seen as a programmable version of
Coq's \lsti{auto}: it currently lacks the latter's capability of
solving goals by \lsti{reflexivity} as well as the modularity of using
\lsti{Hints} as introduction rules; on the bright side, it is not
restricted to depth-first search, since it follows the dictate of the
given FPC, for example iterative-deepening. Furthermore, the FPC can
provide a \emph{trace} that may be more customable that the one
offered by the \lsti{auto}'s hard-wired \lsti{Debug} facility.

@@

As has been detailed in earlier work on foundational proof
certificates, this richer notion of proof system can provide for
richer proof certificates.  The main differences with what we have
seen before is that the left-hand sides of sequents can now grow
during the proof checking process.  As a result, an indexing mechanism
is invoked so that it is possible for the decide expert to properly
select locally added assumptions instead of just those assumptions
accepted globally.

*** To illustrate this approach to elaborating externally provided proof
structures into terms the Coq kernel can check, we consider here
examples take from the file \verb+coq_fpc.v+, which is part of the
repository \url{https://github.com/proofcert/fpc-elpi/}.
Figure~\ref{fig:debruijn} presents the specific FPC that defines the
proof semantics of using the De Bruijn notions for untyped
$\lambda$-terms as proof evidence of a proof in intuitionistic
propositional logic.  That FPC definition, along with a proof checker
capable of synthesizing Coq proof terms during proof checker, makes it
possible to use the term denoting the untyped $\lambda$-term
$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$.

\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
                                                    [lambda (apply 1 [])])])]))
\end{lstlisting}

@@

@@

% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog CIC

>>>>

In this section, we introduce the main ideas from \emph{focused proof
  system}, \emph{foundational proof certificates}, and the
\emph{Coq-Elpi} plugin.

@@

% \todo{Figure now uses Dale kernel-v2 --- needs further abstraction
% and explanation, see appendix}

A \emph{Horn clause} is a formula of the form $\forall \bar
x_1. A_1\supset \forall \bar x_2. A_2\supset \forall \bar
x_n. A_n\supset A_0$ where $\forall\bar x_i$ denote a list of
universal quantifiers ($i\in\{1,\ldots,n\}$) and $A_0,\ldots,A_n$ are
atomic formulas.  It is well known that the following set of sequent
calculus proof rules are complete for both classical and
intuitionistic logic when one is attempting to prove that a given
atomic formula $A$ is provable from a give set $\Pscr$ of Horn
clauses. 
\[
  \infer[decide]{\Pscr\vdash A}
                {D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
  \qquad
  \infer[init]{\Pscr\Downarrow A\vdash A}{}
\]
\[
  \infer[\forall L]{\Pscr\Downarrow\forall x. D\vdash A}
        {\Pscr\Downarrow D[t/x]\vdash A}
  \qquad
  \infer[\supset L]{\Pscr\Downarrow B\supset D\vdash A}
        {\Pscr\vdash B\qquad \Pscr\Downarrow D\vdash A}
\]
Here, we use two different styles of sequents.  The sequent
$\Pscr\vdash A$ is the usual sequent which we generally use as the end
sequent (the conclusion) of a proof.  The sequent $\Pscr\Downarrow
D\vdash A$ is a \emph{focused} sequent in which the formula $D$ is the
\emph{focus} of the sequent.  The two left introduction rules and the
initial rule can only be applied to the focused formula.  This latter
point is in contrast to Gentzen's sequent calculus where these rules
can involve \emph{any} formula on the left of the $\vdash$.  The fact
that this proof system is complete for both classical and
intuitionistic logic (when restricted to the Horn clause fragment)
follows from rather simple considerations of Horn clauses
\cite{nadathur90jacm} and from the completeness of LJT proofs
\cite{Herbelin94}.  The use of the term \emph{focus} comes from
Andreoli's proof system for linear logic \cite{andreoli92jlc}.

@@

Figure~\ref{fig:augmented} contains these proof rules and extends them
in the following fours ways so that it more appropriately fits into
the Coq proof system.
\begin{enumerate}
  \item Instead of having separate connectives for $\forall$ and
    $\supset$, we have the dependent product connective $(x:A)D$. 
  \item We have incorporated both \emph{proof certificates} (using the
    schematic variable $\XXi$) along with expert predicates
    (predicates with the $e$ subscript).
  \item The inference rules are also annotated by terms structures that
    can be give directly to the Coq kernel for checking.
  \item We have added various premises which are responsible to
    pulling information across the Elpi API with Coq.
\end{enumerate}
The proof system described in Figure~\ref{fig:kernel} corresponds to
the Calculus of Inductive Constructions in which all inductive
definitions are limited to the Horn clause.
%
This proof system is inspired by the calculus for proof search in
Pure Type Systems introduced in \cite{LengrandDM06}, based in turn on
ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similar to that calculus, we have a
term language that includes terms and lists of terms, and two distinct
typing judgments for the two categories.
%
In our case, the \lP goal formula \lsti{(check Xi (go B T))}
corresponds to $\XXi\vdash B \colon T$ and the formula
%
\lsti{(check Xi (bc D B L))} corresponds to $\XXi\Downarrow L\colon
D\vdash B$.
%
This style of encode coincides with the idea behind the \emph{spine
  calculus}~\cite{Cervesato97tr}.  The main novelty of our proof
system here is that proof terms and proof certificates are used
simultaneously in all inference rules.

@@

The \emph{unfold} rule is the equivalent in our simplified setting of
a \emph{decide} rule in an explicitly focused system
\cite{liang09tcs}.  From a logic programming viewpoint, given an atom,
the rule selects a clause on which to back-chain on: we look-up the
constructors from an \lsti{Inductive} definition from the global
environment, one that matches the atom we wish to backchain on and then
call the latter judgment. The rules for backchaining are standard,
corresponding to the \emph{init} and the left introduction rules for
$\forall$ and $\supset$.

@@

In Figure~\ref{fig:augmented}  each
inference rule is augmented with an additional premise involving an
\emph{expert predicate}, a certificate $\Xi$, and possibly continuations of
certificates ($\Xi'$, $\Xi_1$, $\Xi_2$) with extracted information from
These notions are taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
\todo{MM: Talk about randomized FPC?}
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:augmented} and~\ref{fig:kernel}:
that is, the exact set of constructors for the  type of certificates as
well as the specification of the expert predicates listed \emph{ibidem}.
%
The top of Figure~\ref{fig:resources} displays two FPCs,
both of which can be used to describe proofs where we bound
the dimension of  a proof.
%
For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) B)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC can be used to bound the total number of instances of
unfoldings in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.

@@

Thanks to the Coq-Elpi interface, in particular to the ``main''
procedure \lsti{solve}, we can package the \lP's code for the checker
as a tactic that can be called as any other tactic in a Coq script.

@@

The glue code between Coq-Elpi and the implementation of our calculus is
straightforward: the goal is given to us as a quadruple of a context, an
evar, a type and a list of extra information. In addition, the certificate is
supplied by the user when calling the tactic (it simply is an integer in this
case). We just need to call \lsti{check} on the goal's type, together with
the certificate, in order to obtain a reconstructed proof term. We avoid
calling the reconstruction directly on the evar because Coq-Elpi ensures that
evars manipulated by \lP are well-typed at all times, and we cannot guarantee
that since we work with partially reconstructed term; getting around this is
simply a matter of assigning the reconstructed term to the evar right
afterwards.
The following example shows how we can use the above tactic to do
FPC-driven logic programming in Coq and return a Coq proof-term:
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
eexists.
elpi  dprolog height 10.
Qed.
Print i1.
ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)) : exists R : list nat, insert 2 [0; 1] R
\end{lstlisting}

The \lsti{dprolog} tactic can be seen as a programmable version of
Coq's \lsti{auto}: it currently lacks the latter's capability of
solving goals by \lsti{reflexivity} as well as the modularity of using
\lsti{Hints} as introduction rules; on the bright side, it is not
restricted to depth-first search, since it follows the dictate of the
given FPC, for example iterative-deepening. Furthermore, the FPC can
provide a \emph{trace} that may be more customized than the one
offered by the \lsti{auto}'s hard-wired \lsti{Debug} facility.

@@

As has been detailed in earlier work on foundational proof
certificates, this richer notion of proof system can provide for
richer proof certificates.  The main differences with what we have
seen before is that the left-hand sides of sequents can now grow
during the proof checking process.  As a result, an indexing mechanism
is used so that it is possible for the decide expert to properly
select locally added assumptions instead of just those assumptions
accepted globally.

\newpage

To illustrate a proof certificate format that uses this richer logic
and proof system, consider the simple \lP code in
Figure~\ref{fig:debruijn}.  Using the constants provided in that
figure, the untyped $\lambda$-term
$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$ can be
encoded as the following Elpi term of type \lsti{deb}.
%
\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
                                                    [lambda (apply 1 [])])])]))
\end{lstlisting}

 presents the specific FPC that defines the
proof semantics of using the De Bruijn notions for untyped
$\lambda$-terms as proof evidence of a proof in intuitionistic
propositional logic.

This FPC can be interpreted by the proof
certificate checker in the file \verb+ljf-dep.mod+ of the repository
\url{https://github.com/proofcert/fpc-elpi/}.
We note that this proof certificate checker is also able to
simulataneously synthesize a Coq proof term while checking (see
Section~\ref{}) all the while remaining less than 90 lines of Elpi
code.

  That FPC definition, along with a proof checker
capable of synthesizing Coq proof terms during proof checker, makes it
possible to use the term denoting the

@@

\newpage

@@

% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog CIC init lp
% LocalWords:  Andreoli's basicstyle lprolog sigs Coq's sortedness Ev
% LocalWords:  backchain gref indt indc coq dest unfoldings str fpc
% LocalWords:  evar nat forall ys elpi Qed
